---
title: "Zendro from R"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

This is an breve documentation about how to use Zendro API from R. 
We will explain with examples how to fetch and use data stored in a Zendro server. 
The process we will follow goes from fetching data to getting that data in a table form. 

## REQUIREMENTS
In order to be able to comunicate with the server we need an R package that allow us to work with HTTP verbs (GET(), POST(), etc).
For this purpose we will use `httr` package.
We will also need other libraries for manipulating json data and tables. 
NOTE: Be aware that you might need to install the correspondant packages.

```{r}
library(httr)         # for http request
library(jsonlite)     # for manipulating json data
library(data.table)   # for table manipulation
library(rlist)        # for table manipulation
library(dplyr)        # for table manipulation
```


## QUERIES EXAMPLES

It's possible to perform any graphql query or mutation by using a POST request. 
Let's start with the most simple example where we will be fetching all `accession_id` and `collectors_name` attributes from `accessions` from a given zendro server served in the url `http://localhost:3001/graphql`.
Due to efficiency reasons it is necessary to provide a pagination limit in the query.


```{r}
url <- "http://localhost:3001/graphql"                                 # set the zendro server url  
accessions_query <- ' 
{
  accessions(pagination:{limit: 6}){
		accession_id
		collectors_name
  }
}
'                                                                     # write the query as a string
result <- POST(url, body = list(query=accessions_query))              # fetch data
```

The `result` that we are getting is the `http` response. We still need to extract the data in order to be able to manipulate it. 
If everything went well, the `http` response will contain an attribute `data` which will itself contain an attribute named as the query, in this case `accessions`.

```{r}
jsonResult <- content(result, as = "text") 
readableText <- fromJSON(jsonResult)
readableText$data$accessions
```

With the above code we were able to visualize the data fetched from zendro server.
Next we will put this data in a table form in order to be able to manipulate the data.

```{r}
dataTable <- lapply(readableText[[1]], as.data.table) 
dataTable$accessions
```